Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-29 paper.tex

\subsection{Interoperability: SAIGE proof-of-concept}
To demonstrate the potential for integrating VCF Zarr into existing utilities,
we developed a proof-of-concept extension of SAIGE~\cite{zhou2018efficiently,
zhou2020scalable}, a popular package for performing genome wide
association tests. Using the Tensorstore C++ library~\cite{tensorstore}
we added support for VCF Zarr via an additional backend class [some details
about the implementation].

To test the performance of this implementation we compared the time required
to perform an association test using the simulated dataset of 10,000 samples
from Fig~\ref{fig-data-storage}. We generated a phenotype using
\texttt{tstrait}~\citep{tagami2024tstrait}.

@@

The VCF Zarr specification is available on GitHub at
\url{https://github.com/sgkit-dev/vcf-zarr-spec/}.
All source code for running benchmarks, analyses and creating
plots in this article is available at
\url{https://github.com/sgkit-dev/vcf-zarr-publication}.
Vcf2zarr is freely available under the terms of the Apache 2.0
license as part of the \texttt{bio2zarr}
suite (\url{https://github.com/sgkit-dev/bio2zarr/})
and can be installed from the Python Package Index
(\url{https://pypi.org/project/bio2zarr/}).
Vcztools (\url{https://github.com/sgkit-dev/vcztools/})
is freely available under the terms of the Apache 2.0
and can be installed from the Python Package Index
(\url{https://pypi.org/project/vcztools/}).

>>>>

\subsection{Interoperability: SAIGE proof-of-concept}
To demonstrate the potential for integrating VCF Zarr into existing
high-performance analysis tools,
we developed a proof-of-concept extension of SAIGE~\cite{zhou2018efficiently,
zhou2020scalable}, a popular package for performing genome wide
association tests. 
Using the TensorStore C++ library~\cite{tensorstore}
we added support for VCF Zarr via an additional backend 
(SAIGE reading data from PLINK, BGEN, VCF, BCF and SAV formats) 
for which the core functionality is defined in a 264 line C++ file.

To assess the performance of this implementation we compared the time required
to perform an association test using the simulated dataset of 10,000 samples
from Fig~\ref{fig-data-storage}. We performed a single-variant association
test based on a phenotype simulated using
\texttt{tstrait}~\citep{tagami2024tstrait},
and measure the overall time required when using the 
equivalent VCF, BCF, SAV and Zarr files.

@@

The VCF Zarr specification is available on GitHub at
\url{https://github.com/sgkit-dev/vcf-zarr-spec/}.
The SAIGE prototype is available at \url{https://github.com/Will-Tyler/SAIGE}
and all other code for running benchmarks, analyses and creating
plots in this article is available at
\url{https://github.com/sgkit-dev/vcf-zarr-publication}.
Vcf2zarr is freely available under the terms of the Apache 2.0
license as part of the \texttt{bio2zarr}
suite (\url{https://github.com/sgkit-dev/bio2zarr/})
and can be installed from the Python Package Index
(\url{https://pypi.org/project/bio2zarr/}).
Vcztools (\url{https://github.com/sgkit-dev/vcztools/})
is freely available under the terms of the Apache 2.0
license and can be installed from the Python Package Index
(\url{https://pypi.org/project/vcztools/}).
